---
title: "Metadata"
author: "Reto Stauffer"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
link-citations: true
vignette: >
    %\VignetteIndexEntry{gsdata: Metadata}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteDepends{gsdata}
    %\VignetteKeywords{GeoSphere, meta data}
    %\VignettePackage{gsdata}
---

```{r, include = FALSE}
library("gsdata")
options("gsdata.cooldown" = 1)
```

Given `mode` and the `resource_id` are known (see `gs_datasets()`)
the function `gs_metadata()` allows to retrieve meta information which provides
which stations and parameters are available for this specific data set.

```{r}
library("gsdata")

# Meta information for historical hourly synop observations
meta <- gs_metadata(mode = "historical", resource_id = "synop-v1-1h")
```

The function returns a named list with a series of information about
the data set (e.g., `title`, `frequency`, `type`, ...) alongside
with two elements `stations` and `parameters` explained in this article
in more detail.

## Station information

Element `stations` is an object of class `r deparse(class(meta$station))`
containing information about available stations for this data set.
This includes the `name` of the station, the geographical location
(`state`, `altitude`, and coordinates) as well as the station `id`
which is required when retrieving data (see `gs_stationdata()`).

In addition, the object provides information about the time period
data is available (`valid_from`, `valid_to`) and whether or not the
station `is_active`.


```{r, fig = TRUE, fig.width = 8, fig.height = 4}
plot(meta$station["altitude"],
     main = "Station location and altitude",
     pch = 19, cex = 1)
plot(meta$station["is_active"],
     main = "Station status (active/inactive)",
     col = ifelse(meta$station$is_active, "#089392", "#CF597E"),
     pch = ifelse(meta$station$is_active, 19, 17),
     cex = 1)
```

## Parameter information

Element `parameters` provides a list of all available parameters for this
data set. Alongside with the `name` which is required to retrieve data
(see `gs_stationdata()`) his `data.frame` also contains a description as well
as the unit of the measurements. This is, however, only available in German.

```{r}
head(meta$parameters, n = 3)
```

Please note that not all stations will provide measurements for all parameters.






