---
title: "Datasets"
author: "Reto Stauffer"
output:
  rmarkdown::html_vignette:
    highlight: monochrome
    toc: true
link-citations: true
vignette: >
    %\VignetteIndexEntry{Package Overview}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteDepends{gsdata}
    %\VignetteKeywords{GeoSphere, datasets}
    %\VignettePackage{gsdata}
---

```{r, include = FALSE}
library("gsdata")
```

TODO(R): Ovrview vignett for R CMD build; the only vignette we'll
include (rest is pkgdown content).

Returns available data sets. Note that the current default is set to
`type = "station"` (see `gs_datasets()`) as this package only allows
for downloading stationdata.

```{r, include = FALSE}
options("gsdata.cooldown" = 1)
```

```{r}
library("gsdata")
(ds <- gs_datasets())
```

The function `gs_datasets()` returns a `data.frame` which includes
a series of information including:

* `type`: data set type (currently defaults to `type = "station"`)
* `mode`: historical or current (latest available observation)
* `resource_id`: ID/name of the dataset or resource
* ... and some more

Most important are `type`, `mode` and `resource_id` which allow
to download additional meta data (see `gs_metadata()`) as well
as the data itself (see `gs_stationdata()`).


