---
title: "gsdata: Interface to the GeoSphere Austria DataHub API (Data Access)"
author: "Reto Stauffer"
output:
  rmarkdown::html_vignette:
    highlight: monochrome
    toc: true
link-citations: true
vignette: >
    %\VignetteIndexEntry{gsdata: Interface to the GeoSphere Austria DataHub API (Data Access)}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteDepends{gsdata}
    %\VignetteKeywords{GeoSphere, datasets}
    %\VignettePackage{gsdata}
---


# Overview

The handling and use of public sector data and information is
standardised across Europe (public sector information; PSI). The aim
is to make public sector data more easily accessible to achieve greater
economic and social benefits.

Since 2021 [GeoSphere Austria](https://geosphere.at/), 
Austria's federal agency for geology, geophysics, climatology and meteorology,
started granting access via their [data hub API](https://data.hub.geosphere.at/).
In early 2024 the API does not only allow to access station data (high quality
in-situ measurements from weather stations) but also spatial data including
high-resolution near-real time analysis, spatial climatologies, but also high-resolution
weather forecasts.

The `gsdata` package (GeoSphere data) provides an easy to use interface to request
and process these data sets. In version `0.0.8` only station data can be processed,
but functionality for gridded data (spatial data) will be added soon.

More detailed overviews and examples are provided in the
[online documentation](https://retostauffer.github.io/gsdata/) of the _gsdata_ package:

* [Datasets](https://retostauffer.github.io/gsdata/articles/datasets.html)
* [Metadata](https://retostauffer.github.io/gsdata/articles/metadata.html)
* [Stationdata](https://retostauffer.github.io/gsdata/articles/stationdata.html)
* [Changelog](https://retostauffer.github.io/gsdata/news/index.html)

# Installation

The stable release of `gsdata` is hosted on the Comprehensive R Archive Network
(CRAN) at <https://CRAN.R-project.org/packge=gsdata> and can be installed via

```{r, eval = FALSE}
install.packages("gsdata")
```

Note that it depends on [`sf`](https://CRAN.R-project.org/package=sf),
thus additional installation of GDAL, GEOS, PROJ and sqlite3 might be needed
(see [`sf` system requirements](https://CRAN.R-project.org/package=sf)).
The development version of `gsdata` is hosted on
[github](https://github.com/retostauffer/gsdata) from where the current
development version can be installed using

```{r, eval = FALSE}
library("remotes")
install_github("retostauffer/gsdata")
```

```{r, include = FALSE}
# Setting cooldown (no more than 1 request every X seconds)
options("gsdata.cooldown" = 0.25)
suppressPackageStartupMessages(library("gsdata"))
```

# Available data sets

The function `gs_datasets()` retrieves information about all available
data sets. By default `type = "station"`, `mode = NULL` is used, thus only
returning information about station data datasets. This can be changed
if needed (see `?gs_dataset`, or in the
[online documentation](https://retostauffer.github.io/gsdata/articles/datasets.html)).

```{r}
library("gsdata")
ds <- gs_datasets()
head(ds, n = 3)
```

```{r}
ds <- gs_datasets(type = "grid", mode = "historical")
head(ds, n = 3)
```

# Metadata

Function `gs_metadata()` allows to retrieve meta information about a specific data
set.


```{r}
meta <- gs_metadata(type = "station", mode = "historical",
                    resource_id = "tawes-v1-10min", expert = TRUE)
names(meta)
```

The return differs for different types of data. If `type = "station"` the
return of `gs_metadata()` contains e.g., information about the time period data
is available (`$start_time`, `$end_time`) but also a data.frame with all available
parameters (`$parameters`; not all parameters are available for all stations).

```{r}
# parameters: data.frame with name, description and unit (DE)
head(meta$parameters, n = 2)
```

In addition a list of available stations is returned (`$stations`)
which contains the station identifier (`id`) required to request
data via the API.

```{r, fig = TRUE, fig.width = 6, fig.height = 3.5, out.width = "100%"}
# stations: a simple feature (sf) data.frame
subset(meta$stations, grepl("INNSBRUCK", name))
plot(meta$stations["altitude"], pch = 19,
     main = "\nLocation and altitude of all available stations")
```

This grants easy access to the data sets meta information, check for available
stations and parameters as well as the `id` of the station which is required
for accessing observational data (next chapter). Note that the station `id`
may vary between different data sets!


# Station data

Function `gs_stationdata()` allows to retrieve data of type `"station"` (measurements
at station level). The following example retrieves temperature (`"TL"`) and
dew point temperature (`"TP"`) observations for January 1, 2024 of two 
automated weather stations (tawes); namely
station Innsbruck airport (`11121`) and Innsbruck University (`11320`).

```{r}
tawes <- gs_stationdata(mode        = "historical",
                        resource_id = "tawes-v1-10min",
                        start       = "2024-01-01",
                        end         = "2024-01-02",
                        parameters  = c("TL", "TP"),
                        station_ids = c(11121, 11320),
                        expert      = TRUE)
names(tawes)
```

`gs_stationdata()` either returns a single `zoo` time series object
(precisely an object of class `c("gs_stationdata", "zoo")`) if data for
one single station is requested, or a list of objects for each of the
stations as in the example above.

```{r, fig = TRUE, fig.width = 6, fig.height = 3.5, out.width = "100%"}
plot(tawes[["11121"]], screen = 1, col = c("red", "limegreen"), lwd = 2)
plot(tawes[["11320"]], screen = 1, col = c("red", "limegreen"), lwd = 2)
```





