---
title: "Datasets"
author: "Reto Stauffer"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
link-citations: true
vignette: >
  %\VignetteIndexEntry{annex: Annex86 Data Analysis Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{Formula}
  %\VignetteKeywords{Annex86}
  %\VignettePackage{annex}
---

```{r setup, include = FALSE}
suppressPackageStartupMessages(library('zoo'))
library("gsdata")
```

This article contains a series of examples on how to retrieve
station observations for different data sets. There is a limit
on how many data can be retrieved via the API in one single call
which is taken care by the function `gs_stationdata()`. If the
number of expected data exceeds the limit, the function will
automatically split thre request in multiple calls. Some
information about that is shown when the argument `verbose` is
set to `TRUE` (`FALSE` by default).

More information about the data sets can be found on the
[Geosphere Austria data hub](https://data.hub.zamg.ac.at/group/stationsdaten).


# Hourly synop observations

## Synop, 1 station, 3 parameters

* Dataset: historical hourly synop observations
* Ten consecutive days 2020
* Parameters: three parameters only
* Stations: single station, `11330` Mayrhofen

```{r data_hist_synop}
mayrhofen <- gs_stationdata(mode        = "historical",
                            resource_id = "synop-v1-1h",
                            start       = "2022-01-01",
                            end         = "2022-01-11",
                            parameters  = c("T", "Td", "ff"),
                            station_ids = 11330)

# zoo object
head(mayrhofen)

# simple zoo plot
plot(mayrhofen,
     screen = c(1, 1, 2),
     col    = c(2, 3, 4),
     ylab   = c("air temperature [C]\nwelt bulb temperature [C]",
                "mean wind speed [m/s]"))
```

As `expert = TRUE` (default) the function will internally download the meta information
calling `gs_metadata()` to check if both arguments provided `station_ids` and `parameters`
are valid (i.e., the station and the parameters are valid for this specific data set).

This can be disabled using `expert = FALSE` which prevents these internal checks and
save some time - but may result in API request errors if the station(s) or parameter(s)
do not exist.

One can, of course, get the meta information for the station and parameters
if needed (see `gs_metdatata()` for more details):


```{r meta_hist_synop}
meta <- gs_metadata(mode = "historical", resource_id = "synop-v1-1h")
subset(meta$stations, id == 11330)
subset(meta$parameters, name %in% c("T", "Td", "ff"))
```

### Synop, multiple stations, all parameters

* Dataset: historical hourly synop observations
* Ten consecutive days 2020
* Parameters: all available parameters (uses `parameters = NULL`; default)
* Stations: three different stations

The function allows to retrieve data for multiple stations at once and to
download all avialable parameters by not specifying `parameters` (uses `parameters = NULL`).
This will return all avialable data for each of the stations in a named list, where
the names of the elements correspond to the station `id`.

```{r data_hist_synop_innsbruck, warnings = TRUE}
x <- gs_stationdata(mode        = "historical",
                    resource_id = "synop-v1-1h",
                    start       = "2020-01-01",
                    end         = "2020-01-11",
                    station_ids = c(11330, 11328, 11120),
                    expert      = TRUE)
class(x)
names(x)
```

`expert = TRUE` skips the internal check whether or not the `station_ids` given
are valid (i.e., if they even exist for this data set).

Note that station `11328` does not provide any data for the time period specified resulting
in a warning and an will return a missing value (`NA`).

```{r}
x[["11328"]]
is.na(x[["11328"]])
```

The other two elements of the list again contain objects of class 
`r deparse(class(x[[1]]))`:

```{r}
lapply(x[c("11330", "11120")], class)
lapply(x[c("11330", "11120")], head, n = 2)
```

The API, however, provides different data for the other two stations as these
two stations do not provide the data for the very same parameters.  Parameters
not providing any data will not be included in the return.

```{r, plot = TRUE, fig.width = 12, fig.height = 10}
lapply(x[c("11330", "11120")], dim)
# Default plots
plot(x[["11330"]], nc = 3, col = "steelblue", type = "o", pch = 4)
plot(x[["11120"]], nc = 3, col = "steelblue", type = "o", pch = 4)
```


# TAWES data

TAWES stands for "teilautomatische Wetterstation" and provides data
on a 10 minute temporal resolution. Requesting data for TAWES stations
works the very same way as for synop stations as shown above, however,
the set of avialble stations and parameters is different (names of the
parameters differ).

To get information about all available stations and parameters call:

```{r data_hist_tawes}
meta <- gs_metadata("historical", "tawes-v1-10min")
```

... and check the vignette [Metadata](articles/metadata.html) for more
information on the return.


# Klima

## Daily climate records, 1 station, 3 parameters

* Dataset: historical climate recoreds; daily temporal resolution
* Summer 2020 till the end of 2022 (2.5 years)
* Parameters: three distinct parameters (precipitation and snow height)
* Stations: single station, `8807` Achenkirch

```{r, meta_data_hist_klima_achenkirch, plot = TRUE, fig.width = 10, fig.height = 6}
meta <- gs_metadata("historical", "klima-v1-1d")
achenkirch <- gs_stationdata(mode        = "historical",
                             resource_id = "klima-v1-1d",
                             start       = "2020-06-01",
                             end         = "2022-12-31",
                             parameters  = c("nied", "nied07", "schnee"),
                             station_ids = 8807,
                             expert      = TRUE)
head(achenkirch)
plot(achenkirch, type = "h")
```


## 10min climate records, 1 station, 3 parameters

* Dataset: historical climate recoreds; daily temporal resolution
* Summer 2020 till the end of 2022 (2.5 years)
* Parameters: three distinct parameters (precipitation and snow height)
* Stations: single station, `8807` Achenkirch

```{r meta_data_hist_klima_uibk, plot = TRUE, fig.width = 10, fig.height = 6}
meta <- gs_metadata("historical", "klima-v1-10min")
uibk <- gs_stationdata(mode        = "historical",
                       resource_id = "klima-v1-10min",
                       start       = "2010-11-01",
                       end         = "2011-02-01",
                       parameters  = c("TL", "FFAM", "FFX"),
                       station_ids = 11803,
                       verbose = TRUE,
                       expert      = TRUE)
plot(uibk,
     screens = c(1, 2, 2),
     col = c(2, 4, 8),
     ylab = c("temperature", "mean wind\nand gusts"))
```

## TAWES, 1 station, 4 

* Dataset: historical TAWES observations, 10min temporal resolution
* Past 30 consecutive days (based on system date)
* Parameters: four predefined parameters
* Stations: single station, `11121` Innsbruck Airport

**WARNING:** `resource_id = "tawes-v1-10min"` is not qualiy controlled
and seems to provide a limited amount of data (not going too far back in time).
Consider using `resource_id = "klima-v1-10min"` instead (see above) which
provides long-term historical quality-controlled data. Station IDs (and
available parameters) differ, tough.

```{r data_hist_tawes_ibkap, plot = TRUE, fig.width = 10, fig.height = 6}
ibkap <-  gs_stationdata(mode        = "historical",
                         resource_id = "tawes-v1-10min",
                         start       = Sys.Date() - 30,    # Today - 30 days (system time)
                         end         = Sys.Date(),         # Today (system time)
                         parameters  = c("TL", "TP", "FFAM", "FFX"),
                         station_ids = 11121,
                         expert      = TRUE)
plot(ibkap,
     screens = c(1, 1, 2, 2),
     col = c(2, 3, 4, 8),
     ylab = c("temperature\nand dewpoint", "mean wind\nand gusts"))
```


# Histalp

## Histalp annual data, 1 station, 2 parameters

* Dataset: historical histalp records; annual observations
* 1845 to 2022
* Parameters: precipitation and temperature (annual)
* Stations: single station, `23` Bregenz

**NOTE:** Not executed; requires login; currently throws an error.

```{r meta_data_hist_histalp_login, plot = TRUE, fig.width = 10, fig.height = 6, error = TRUE}
meta <- gs_metadata("historical", "histalp-v1-1y")
bregenz <- gs_stationdata(mode        = "historical",
                          resource_id = "histalp-v1-1y",
                          start       = "1854-01-01",
                          end         = "2022-01-01",
                          parameters  = c("R01", "T01"),
                          station_ids = 23,
                          expert      = TRUE)

##plot(bregenz, col = c(4, 2))
```




